The Limits of Loyalty: Postmodernism, Polarization, and Military Professionalism  (Prospectus)

### 2020-09-26 code/prospectusv1.Rmd
commit

  _This dissertation seeks to explore the relationship of postmodernism and political polarization and their impact on the US military profession. The fundamental question that I seek to answer is, "how is rising political polarization shaping the military as a profession in the United States, and what are the implications of these changes, both in terms of national security and the future of the military profession?" In answering this question, I develop a theory that, at its core, argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession. I further argue that the strain of loyalty is especially problematic for military, as it is a profession that by its nature holds mutiple loyalties simultaneously: to the Constitution, to the President as Commander-in-Chief, to Congress, and to the American People. Balancing these competing loyalties becomes more difficult, I argue, as political polarization worsens, because of the fundamental values - not mere policy differences - that are at the root of many polarized political battles today. In the dissertation, I seek to employ a mixed-methods approach that incorporates several case studies, as well as quantitative analysis, to support my theory. This dissertation will contribute to the fields of American politics, international security, and civil-military relations by further illuminating the mechanisms through which political polarization is impacting the military institution. Specifically, it will provide a contrarian view to recent civil-military relations scholars who have argued that the traditional concept of military professionalism should be significantly redesigned, if not abandoned entirely. This dissertation will argue the opposite: that what is needed most in politically polarizing times, such as those the US is in right now, is a return to and embrace of traditional conceptions of military professionalism - an effort that will require deliberate action by not just military leaders, but elected civilian leaders as well._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

Thankfully, a small but growing number of civil-military relations scholars have looked at aspects of this question in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. These scholars have made important findings that speak to the state of current American civil-military relations. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

In their own right, these scholarly developments are interesting and important, but another debate in the civil-military relations field has recently emerged, or to be more correct, reawakened. At the center of this debate is the degree to which the concept of _military professionalism_ should be altered, if not discarded altogether [@brooks_paradoxes_2020], by the collective civilian-military leadership, in part due to concerns about the tense domestic political environment that exists in contemporary America. The central question in this debate is whether the concept of _military professionalism_ actually accomplishes its intended goals: to simultaneously achieve military security from external threats, and to maintain civilian control of the military -- an argument developed in Samuel Huntington's 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_.

In this seminal work, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims constrains the military in that the military can never lose certain aspects such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs a strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society.

Huntington asserts that the key to resolving the tension generated by these two opposing forces - to win on the battlefield and to maintain itself as a liberal democracy - is to adopt a scheme of "objective civilian control" whereby political leaders maximize the professional autonomy of the military by establishing the ends and goals of policy, but allowing military leaders to develop and implement the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. The creation and maintenance of largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military would win on the battlefield, but also stay out of politics and thus, untainted by partisan interests and concerns.

@@

## The Concept and Norm of Organizational Loyalty

An important value among militaries, especially professional militaries, is loyalty. Indeed, the US Army's first "Army Value" is "loyalty," followed by six other traits that in totality, spell out the acronym, l-d-r-s-h-i-p, or "leadership": "integrity," "duty," "respect," selfless service," "honor," integrity," and "personal courage" [need cite here]. The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [need cite here for ADP 6-22, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [need cite here for ADP 6-22.]

@@

One area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity of the military, both as an American institution, but also as a profession. Exploring this relationship will, I believe, help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it would, in the sense that the military, at least in America, is comprised of people from all walks of life. On the other hand, one might expect that the unique purpose and "professional" nature of the military might insulate it from some of the impacts of polarized politics. I am especially interested, to be more specific, in if and how political polarization impacts the concept of organizational loyalty in the military, particularly because one of the hallmarks of the military's professional nature is that it ought to be apolitical and non-partisan. Yet in an environment where the ability to be apolitical and non-partisan is becoming more and more difficult, due to rising political polarization, it is important that scholars and policymakers understand the mechanisms through which the military is being shaped, if there any, resulting in changes to its professional character.

To be clear, scholars have long studied how domestic politics impact the military. To be sure, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My personal motivation in choosing the intended topic, however, stems from what I believe are changes in the type of civil-military disagreement that have surfaced in recent years. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, way outside of appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] More interestingly, these conflicts are occurring not in the context of major war, but the opposite: a period in which the levels of US Troops deployed in active combat zones is less than it has been in recent years. I further believe that while such developments are catching the attention of civil-military relations scholars, most of the work done on this topic is impressionistic [@golby_jim_keep_2020] as opposed to systematic. In other words, civil-military relations scholars have identified that polarized partisan politics serve as a risk for the military, but none to my knowledge have have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character.

My intended dissertation will help answer several related questions, including: if personal and partisan loyalties are changing as a result of political polarization [need cite], is this having an impact within the military, i.e. is the concept of organizational loyalty changing as a result of polarization? Are these changes in organizational loyalty having an impact on the military's professional identity and character? Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? in the military changing as a result of polarized domestic politics? If so, through what mechanisms, and how much? Is this problematic for the military profession, and why? The answers to these questions, I believe, will help inform the broader field of international relations on the broader and much more holistic issue of how changes in political polarization impact the military's ability to defend the nation and achieve the goals of American foreign policy.

# The Analytical Framework
## Polarization and Military Professionalism

how increasing domestic political polarization impacts conceptions of loyalty in military professions, and to examine the implications of these changes in loyalty on broader security outcomes. For at least the past 75 years[need footnote here], the US military has prided itself on being not merely an organization charged with the defense of the nation, but a profession. That is, on being an institution marked by several unique traits, including: providing a unique service that requires special skills (the application and management of violence); establishing and maintaining standards of behavior and performance, as well as policing itself when members of that profession fail to uphold these standards; and building and maintaining trust between itself, the profession, with the client whom it serves (the American people) [insert source here]. The US military, further, is both by statute and by normative standards an apolitical and non-partisan institution with multiple, simultaneous loyalties: to the Constitution of the United States; to the President of the United States as the Commander-in-Chief; and to Congress. But what happens if and when the objects of the military's multiple loyalties not merely disagree, but become severely divided ideologically? Can such an outcome exist, and if it can, is there empirical proof that such a trend is occurring? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

Figure One, below, shows one possible way to conceptualize the system of problems generated by increasing political polarization and its impact on the military profession, as well as the roles of other structural variables, including the institutional credibility of the military profession, and other various military outcomes. From left to right, this conceptual model suggests that aspects of political polarization are simulataneously driving changes in the institutional credibility of the military as well as changes in the professional character/nature of the military. This is made more complicated, as the model suggests, that the intervening variable of institutional credibility is also driving some changes in the professional character of the military.

In my dissertation, I situate changing conceptions of loyalty as a major subcomponent of the arrow running from political polarization to changes in military professionalism. At the same time, I do think that political polarization is impacting the credibility of the military institution, which in turn may likewise be driving changes in military professionalism. The specific mechanisms through which this is occurring, if at all, however, need to be developed, proposed, and tested.

```{r figure one, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    MO ~ MP,
    MP ~ PP + IC,
    IC~PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MO", 1, .5,
      "MP", .7, .5,
      "IC", .3, .2,
      "PP", .1,.5,
    ))

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Institutional\n Credibility")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

At this stage in the dissertation phase, I do not know how polarization is impacting conceptions of loyalty, or even if it is, for sure. I strongly suspect that it is, however, for a couple of reasons. First, the military is, I believe, susceptible to charges of failing to uphold the ideal of loyalty based on the nature of the organization in that it ultimately is charged to enact violence on behalf of the state. Harold Lasswell said of military officers that their skill is the "management of violence" [need quote from Huntington]. As the bonds of loyalty forged on the battlefield are strong and different in ways than loyalty forged in other environments, I suspect that the military and its leaders will take very seriously any accusations that the corporate entity or individual officers are failing to be loyal to a number of ideals. In the Army, for example, the first "Army Value," as it is called by the militar service,  which is summarized as the acronym, "LDRSHIP" (short for "leadership"), is the word "loyalty." [need cite for the Army values here].

@@

To complete my dissertation, I plan on incorporating a mixed-methods approach that relies on first, proposing a theory for how political polarization impacts and affects the overarching structure of loyalty that military officers - and the military as an institution - face. I then seek to test that theory using both empirical and quantitative data, as well as a series of qualitative case studies. I plan to conclude by discussing the implications of my findings. My dissertatino will be a book style projecte composed of at least seven chapters.

>>>>

  _This dissertation seeks to explore the relationship of postmodernism and political polarization and their impact on the US military profession. The fundamental question that I seek to answer is, "how is rising political polarization shaping the military as a profession in the United States, and what are the implications of these changes, both in terms of national security and the future of the military profession?" In answering this question, I develop a theory that, at its core, argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession. I further argue that the strain of loyalty is especially problematic for military, as it is a profession that by its nature holds multiple loyalties simultaneously: to the Constitution, to the President as Commander-in-Chief, to Congress, and to the American People. Balancing these competing loyalties becomes more difficult, I argue, as political polarization worsens, because of the fundamental values - not mere policy differences - that are at the root of many polarized political battles today. In the dissertation, I seek to employ a mixed-methods approach that incorporates several case studies, as well as quantitative analysis, to support my theory. This dissertation will contribute to the fields of American politics, international security, and civil-military relations by further illuminating the mechanisms through which political polarization is impacting the military institution. Specifically, it will provide a contrarian view to recent civil-military relations scholars who have argued that the traditional concept of military professionalism should be significantly redesigned, if not abandoned entirely. This dissertation will argue the opposite: that what is needed most in politically polarizing times, such as those the US is in right now, is a return to and embrace of traditional conceptions of military professionalism - an effort that will require deliberate action by not just military leaders, but elected civilian leaders as well._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

Thankfully, a small but growing number of civil-military relations scholars have looked at aspects of this question in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. These scholars have made important findings that speak to the state of current American civil-military relations. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military, perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

In their own right, these scholarly developments are interesting and important, but another debate in the civil-military relations field has recently emerged, or to be more correct, reawakened. At the center of this debate is the degree to which the concept of _military professionalism_ should be altered, if not discarded altogether [@brooks_paradoxes_2020]. The central question in this debate is whether the concept of _military professionalism_ actually accomplishes its intended goals: to simultaneously achieve military security from external threats, and to maintain civilian control of the military -- an argument developed in Samuel Huntington's 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_.

In this seminal work, which often serves as a point of departure for many discussions and debates within the field of civil-military relations, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims actually constrains the military in that the military can never become so alienated from society such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. The societal imperative, Huntington claimed, was a further function of two distinct features of the United States. The first was America's liberal ideology [need to confirm this name of ideology], which Huntington saw as fundamentally suspicious of standing military forces, dating back to the Revolutionary War period [need cite]. Moreover, American liberal ideology for Huntington also values the individual extensively. The second feature of the societal imperative, claimed Huntington, was the US constitution, which clearly states that the military will always be controlled by civilians. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs a strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society.

Figure one, below, shows a visual diagram of Huntington's logic in a political science framework (though as @feaver_civil-military_1996 points out, Huntington did not use such language when articulating his concepts). According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that impacts and enables these goals while also resolving the tension generated by the opposing functional and societal imperatives, as described above, as adoption of a model of "objective civilian control" of civil-military relations, whereby political leaders maximize the professional autonomy of the military by establishing the ends and goals of policy, but allow military leaders to develop and implement the military plans to accomplish these goals [@huntington_soldier_1957, 80-85].

In making his argument, Huntington went to great lengths to describe an alternate world in which the pattern of civilian control, the key independent variable impacting military security and civilian supremacy of the military, was not "objective control," but "subjective control" [need cite for Huntington]. In this conception, Huntington saw civilian leaders making the military a political object through a variety of means, including the selection of military leaders sympathetic to partisan political causes [need cite], to the point that there was no longer any distinction between the military and civilian leadership [need cite]. To Huntington, adopting this form of control, and thus erasing what he saw as a necessary distinction between both the military and civilian leadership and society, would endanger American security. By virtue of its distinct mission and functional goals, Huntington asserted that the military needed to operate apart from partisan politics - something it could not do if a pattern of subjective control were to be adopted and implemented - in order to maintain military security.

```{r figure one, include=FALSE, echo=FALSE}
huntington_dag <- dagify(
    MSCS ~ CC,
    CC ~ FI + SI,
    SI~I + S,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MSCS", 1, .5,
      "CC", .7, .5,
      "FI", .5, .6,
      "SI", .5, .4,
      "I", .3, .3,
      "S",.7, .3,
    ))

huntington_dag %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text_repel(aes(label=name), show.legend=FALSE)+
  theme_dag()
```

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One: Samuel Huntington's Theory\n Of Civilan Control of the Military", subtitle="As espoused in The Soldier and The State, 1957")

@@

As an extension of the critiques of Huntington, there has also been renewed focus and attention among civil-military relations scholars on numerous professional normative standards of the military. One of the most important yet difficult debates is an attempt by scholars to define where appropriate political behavior ends and partisan activity begins [@owens_military_2015; @golby_beyond_2015]. In the Huntingtonian framework, this is essentially the opposite side of the debate over the same coin in attempting to define for civilians what constitutes meddling in military affairs. One of the reasons this debate has become even more difficult, I believe, is because as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms.

## Military Professions and Organizational Loyalty

Loyalty features prominently in the international relations and civil-military relations literatures, often in a comparative sense and in the context regime exclusion and a host of related phenomena, including military coups [need cite], civil wars [@mclauchlin_loyalty_2018], and the difficulties that multi-ethnic states face in maintaining a professional military [@peled_question_1998]. This literature deals with a host of important issues that stem from problems associated with ethnic identification and loyalty, either to a regime, cause, or leader. Though the US context is very different from the contexts usually explored in this literature, its insights may help me identify related dynamics and/or mechanisms that perhaps are unfolding in the US: how are changes in partisan or ideological identification (instead of ethnic) as a result of political polarization impacting conceptions of loyalty within the military. Because appropriate to explore partisan or ideological (instead of ethnic) identification stemming from However, this literature often assumes a definition of loyalty and rarely develops how the reader should think of loyalty in the first place. For this fundamental starting point of defining loyalty, the fields of ethics, management, and sociology are helpful.

Loyalty has been conceptualized as both a virtue [@coleman_problems_2009] and as a moral emotion [@connor_military_2019]. Though the differences between the two approaches are important, both assert that loyalty is something felt by an individual and subsequently reflected in action, such as, in the military context, "the desire to protect each other from harm and to defend each other from attack" [@coleman_problems_2009, p. 109]. Oldenquist, a philosopher, describes loyalty as a devotion or attachment that an individual has to someone or something specific such that one refers to the object of loyalty as "mine" and is inclined to feel "pride when it [the object of one's loyalty] prospers, shame when it declines, and anger or indignation when it is harmed" [@oldenquist_loyalties_1982, p. 175, brackets mine].

Loyalty further generates a willingness in the individual to assume "obligations" [@oldenquist_loyalties_1982, p. 175] to the object of one's loyalty, and these obligations can take multiple forms, such as a willingness of the individual to endure hardships and make sacrifices [@fielder_organizational_1992]. Scholars agree that individuals can hold multiple loyalties simultaneously, and recognize that this often can and does cause problems [@coleman_problems_2009]. For organizations, and by extension, professions, the concept of loyalty is also very important. Conner et al. state, for example, that "having loyal members is vital to perpetuating the group" [@connor_military_2019].

We also know that the military services in the United States place a premium on loyalty as a trait and concept. Indeed, of the different military services, the US Army in particular has extensively developed, in a normative sense, a framework entitled "The Army Ethic," in which loyalty constitutes a significant part of the Army profession [@us_department_of_the_army_adp_2019]. This ethic, which will be described in detail in later parts of this dissertation, is formally laid out and described in the Army's doctrinal publication entitled, _Army Leadership and the Profession, ADP 6-22_, and holds that as "stewards of the Army Profession" and as "committed professionals," Army leaders "embrace and uphold the Army Values and standards of the profession" [@us_department_of_the_army_adp_2019]. There are a total of seven "Army Values," which together spell the acronym, l-d-r-s-h-i-p, or "leadership," and the first of these values, loyalty, is followed by six others: duty, respect, selfless service, honor, integrity, and personal courage [@us_department_of_the_army_adp_2019, p. 1-9.].  The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [need cite here for ADP 6-22, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [need cite here for ADP 6-22.]

Loyalty is further important in a military setting in a broader and strategic sense, especially when one considers that the US military, both by statute and by normative standards, serves as an apolitical and non-partisan institutions with multiple, simultaneous loyalties. The Constitution of the United States, to which military officers and soldiers alike ascribe allegiance when they take their respective oaths of commissioning and enlistment, deliberately contrives the military as an institution possessing multiple loyalties. One of the most basic forms of these multiple loyalties is captured in the fact that the Constitution empowers the President as "the Commander in Chief of the Army and the Navy of the United States" [@noauthor_us_1787], but also gives Congress an array of tremendous powers and responsibility pertaining to the Armed Forces @[noauthor_us_1787-1]. Indeed, the tensions that inherently exist in American's constitutional framework are explored in a robust literature that examines the multiple ways in which the Congress, the President, and the military interact to implement, but also to influence, American foreign policy [for instance, see @howell_while_2007].

In the context of the US military as an institution, loyalty has rarely been used as a conceptual construct in the political science literature, likely and thankfully because the US military has not yet considered, or been involved, in a coup. Nonetheless, I contend that there is a rich opportunity to study strategic loyalty as a concept, especially in a contemporary context in the United States, where political polarization has increased extensively.

@@

One area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity of the military, both as an American institution and as a profession. Exploring this relationship will, I believe, help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it might, in the sense that the American military is, after all, a microcosm of American society. Therefore, one could expect that if society truly is becoming polarized, then at least to some degree, so is the military, or at least those who volunteer for service and at the point in time that they join. On the other hand, one might reasonably expect that the unique purpose of the military, combined with its long-standing professional ethic, might insulate the military institution from some of the potentially negative impacts of polarized politics. These differing expectations need to be more clearly articulated and explored.

I am specifically interested in the question of if and how political polarization impacts the concept of organizational loyalty in the military. This is an interesting research question because, as noted above, one of the most fundamental normative hallmarks of the military's professional nature is that it ought to be apolitical and non-partisan, and one of the features of militaries in general is their tendency to foster a concept of loyalty among those who serve. Yet a puzzle emerges: within an institution that is normatively apolitical and non-partisan, can it remain so in an environment marked by increasingly polarized domestic politics, where partisan loyalties are a salient feature of the political landscape [need cite here for partisan loyalties].

My intended dissertation will help answer several related questions, including: if personal and partisan loyalties are changing as a result of political polarization [need cite], is this also happening within the military institution, and why? Are partisan/ideological loyalties displacing other loyalty relationships within the military? If so, are these changes in organizational loyalty having an impact on the military's professional identity and character? Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? Is this problematic for the military profession, and why?

The answers to these questions, which I intend to include as part of my dissertation, will help inform scholars and policymakers alike on number of important issues. First, it will help illuminate in a broad sense an important aspect of how changes in political polarization impact American institutions, and in particular, the military. Second, work on this topic will attempt to explore how a fundamental concept and value within the military - loyalty - is both impacted as a result of political polarization but also how loyalty changes may be driving changes in the military's professional character, identity, and nature. In doing so, it will help enrich a scholarly understanding of what loyalty is to begin with, and how it features prominently in the context of the American military.

To be clear, scholars have long studied how domestic politics impact the military. And as scholars have pointed out, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My personal motivation in choosing the intended topic, however, stems from what I believe are changes in the type of civil-military disagreement that have surfaced in recent years. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, way outside of appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] More interestingly, these conflicts are occurring not in the context of major war, but the opposite: a period in which the levels of US Troops deployed in active combat zones is less than it has been in recent years. While I believe that such developments are catching the attention of civil-military relations scholars, most of the work done on this topic is, in my opinion, impressionistic [@golby_jim_keep_2020] as opposed to systematic. In other words, civil-military relations scholars have identified that polarized partisan politics serve as a risk for the military, but none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and how.

# The Analytical Framework - Initial Arguments and Hypotheses

Given all that the existing scholarship says about the topics of political polarization, military professionalism, and concepts of loyalty and identity, the goal in my dissertation is to examine how these concepts are inter-related, and specifically, if and how polarization drives changes in the way that these concepts are inter-related. Further, I want to determine if these changes impact national security, and if so, in what ways?

Figure two below, shows an initial way to link some of the variables that I am interested in exploring throughout my dissertation. From left to right, this conceptual model suggests that aspects of political polarization are simultaneously driving changes in both the institutional credibility of the military (how society views the military), as well as changes in the conceptions of loyalty and identity (within the military). These two variables are, in turn, driving changes in the concept of military professionalism, my main dependent variable, within the military, and I suspect that these changes may, if not yet, potentially impact various military outcomes. By various military outcomes, I mean the very same military outcomes that other scholars do and have already explored, such as effectiveness on the battlefield, i.e winning wars and battles [@reiter_democracies_2002; @biddle_democracy_2004] and achieving strategic effectivness [@brooks_creating_2007; @brooks_shaping_2008]. While I intend to examine this system of variables, my focus will be on the link between increasing political polarization, the concepts of loyalty and identity, and military professionalism, both in order to focus my research but also because there has been more work done recently examining other variables, to include the impact of increasing political polarization on institutional credibility [@robinson_michael_danger_2018].

```{r figure two, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    MO ~ MP,
    MP ~ LI + IC,
    IC~PP,
    LI~PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MO", 1, .5,
      "MP", .7, .5,
      "IC", .3, .3,
      "LI", .3, .7,
      "PP", .1,.5,
    ))

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Institutional\n Credibility",
                     "LI"="Concepts of\n Loyalty and Identity")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

A primary argument is...

how increasing domestic political polarization impacts conceptions of loyalty in military professions, and to examine the implications of these changes in loyalty on broader security outcomes. For at least the past 75 years[need footnote here], the US military has prided itself on being not merely an organization charged with the defense of the nation, but a profession. That is, on being an institution marked by several unique traits, including: providing a unique service that requires special skills (the application and management of violence); establishing and maintaining standards of behavior and performance, as well as policing itself when members of that profession fail to uphold these standards; and building and maintaining trust between itself, the profession, with the client whom it serves (the American people) [insert source here].  But what happens if and when the objects of the military's multiple loyalties not merely disagree, but become severely divided ideologically? Can such an outcome exist, and if it can, is there empirical proof that such a trend is occurring? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

In my dissertation, I situate changing conceptions of loyalty as a major sub component of the arrow running from political polarization to changes in military professionalism. At the same time, I do think that political polarization is impacting the credibility of the military institution, which in turn may likewise be driving changes in military professionalism. The specific mechanisms through which this is occurring, if at all, however, need to be developed, proposed, and tested.

At this stage in the dissertation phase, I do not know how polarization is impacting conceptions of loyalty, or even if it is, for sure. I strongly suspect that it is, however, for a couple of reasons. First, the military is, I believe, susceptible to charges of failing to uphold the ideal of loyalty based on the nature of the organization in that it ultimately is charged to enact violence on behalf of the state. Harold Lasswell said of military officers that their skill is the "management of violence" [need quote from Huntington]. As the bonds of loyalty forged on the battlefield are strong and different in ways than loyalty forged in other environments, I suspect that the military and its leaders will take very seriously any accusations that the corporate entity or individual officers are failing to be loyal to a number of ideals. In the Army, for example, the first "Army Value," as it is called by the military service,  which is summarized as the acronym, "LDRSHIP" (short for "leadership"), is the word "loyalty." [need cite for the Army values here].

@@

To complete my dissertation, I plan on incorporating a mixed-methods approach that relies on first, proposing a theory for how political polarization impacts and affects the overarching structure of loyalty that military officers - and the military as an institution - face. I then seek to test that theory using both empirical and quantitative data, as well as a series of qualitative case studies. I plan to conclude by discussing the implications of my findings. My dissertation will be a book style project composed of at least seven chapters.
